---
title: "Week 4 - Assignment 3 - Quiz - Attempt 1"
author: "Sally Longmore"
date: "02/02/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
source("best.r")
source("rankhospital.r")
source("rankall.r")
```

Question 1. What was the result returned by the following code?

```{r}
best("SC", "heart attack")
```

Quetion 2. What result is returned by the following code?

```{r}
best("NY", "pneumonia")
```

Question 3. What result is returned by the following code?

```{r}
best("AK", "pneumonia")
```

Question 4. What result is returned by the following code?

```{r}
rankhospital("NC", "heart attack", "worst")
```

Question 5. What result is returned by the following code?

```{r}
rankhospital("WA", "heart attack", 7)
```

Question 6. What result is returned by the following code?

```{r}
rankhospital("TX", "pneumonia", 10)
```

Question 7. What result is returned by the following code?

```{r}
rankhospital("NY", "heart attack", 7)
```

Question 8. What result is returned by the following code?

```{r}
r <- rankall("heart attack", 4)
as.character(subset(r, state == "HI")$hospital)
```

Question 9. What result is returned by the following code?

```{r}
r <- rankall("pneumonia", "worst")
as.character(subset(r, state == "NJ")$hospital)
```

Question 10. What result is returned by the following code?

```{r}
r <- rankall("heart failure", 10)
as.character(subset(r, state == "NV")$hospital)
```

