---
title: "Week 2 - R Basics - Dates and Times"
author: "Sally Longmore"
date: "24/01/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

R has a special way of representing dates and times, which can be helpful if you're working with data that shows how something changes over time (i.e. time-series data), or if your data contains some other temporal information, like dates of birth.

Dates are represented by the **Date** class and times are represented by the **POSIXct** and **POSXlt** classes. Internally, dates are stored as the number of days since 1970-01-01 and times are stored either as the number of seconds since 1970-01-01 for **POSIXct** or a list of seconds, minutes, hours etc for **POSIXlt**.

Lets start by using d1 <- Sys.Date() to get the current date and store it in the variable d1. (That's the letter **d** and the number **1**)

```{r}
d1 <- Sys.Date()
```

Use the class() function to confirm that d1 is a Date object.

```{r}
class(d1)
```

We can use the unclass() function to see what d1 looks like internally

```{r}
unclass(d1)
```

That's the exact number of days since 1920-01-01

However, if you print d1 to the console, you will get today's date in yyyy-MM-dd format.

```{r}
d1
```

What if we need a reference to a dat prior to 1970-01-01. Create a variable d2 containing the date 1969-01-01.

```{r}
d2 <- as.Date("1969-01-01")
```

Now use the unclass() again to see what d2 looks like internally

```{r}
unclass(d2)
```

As you may have anticipated, you get a negative number. In this case it is -365, since 1969-01-01 is exactly one calendar year (i.e. 365 days) before 1970-01-01.

Now, let's take a look at how R stores times, you can access the current date and time using the Sys.time() function with no arguments. Do this and store the result as t1.

```{r}
t1 <- Sys.time()
t1
```

An check the class() of t1.

```{r}
class(t1)
```

As mentioned earlier, POSIXct is just one of two ways that R represents time information. (You can ignore the second value above, POSIXt, which functions as a common language between POSIXct and POSIXlt).  

Use unclass() to see what t1 looks like iternally, the (large) number of seconds since the begining of 1970.

```{r}
unclass(t1)
```

By default, Sys.teim() returns an object of class POSIXct, but we can coerce the result to POSIXlt with as.POSIXlt(Sys.time()). Give it a try and store the result in t2.

```{r}
t2 <- as.POSIXlt(Sys.time())
```

Check the class of t2.

```{r}
class(t2)
```

Now view it's contents.

```{r}
t2
```

The printed format of t2 is identical to that of t1. No unclass() t2 to see how it is different internally.

```{r}
unclass(t2)
```

t2, like all POSIXlt objects, is just a list of values that make up the date and time. Use str(unclass(t2)) to have a more compact view.

```{r}
str(unclass(t2))
```

If, for example, we want just the minutes from the time stored in t2, we can access them with t2$min.

```{r}
t1$min
```

Now that we have explored all three types of date and time objects, let's look at a few funcitons that extract useful information from any of these objects -- weekdays(), months() and quaters().

The weekdays() function will return the day of the week from any date or time object. Try it out on d1, which is the date object that contains today's date.

```{r}
weekdays(d1)
```

The months() function also works on any date or time object. Try it on t1, which is the POSIXct object that contains the current time (well, it was the current time at the time you created it).

```{r}
months(t1)
```

The quaters() function returns the quater of the year (Q1-Q4) from any date or time object. Try it on t2, which is the POSIXlt object that contains the time at which you created it.

```{r}
quarters(t2)
```

Often, the dates and times in a dataset will be in a format that R does not recognise. The strptime() function can be helpful in this situation.

strptime() converts character vectors to POSIXlt. IN that sense, it is similar to as.POSIXlt(), except that the input doesn't have to be in a particular format (yyyy-MM-DD).

To see how it works, store the following character string in a variable called t2:

"October 17, 1986 08:24"

```{r}
t3 <- "October 17, 1986 08:24"
```

Now, use strptime(t3, "%B %d, %Y %H:%M") to help R convert our date/time object to a format that it understands. Assign the result to a new variable called t4.

```{r}
t4 <- strptime(t3, "%B %d, %Y %H:%M")
t4
```

That's the format we've come to expect. Now, let's check it's class()

```{r}
class(t4)
```


Finally, there are a number of operations that you can perform on dates and times, including arithmetic operations (+ and -) and comparisons (<, ==, etc.)

The variable t1 contains the time at which you created it. Confirm that some time has passed since you created t1 by using the > operator to compate it to the current time.

```{r}
Sys.time() > t1
```

So, we know that some time has passed, but how much? Try subtracting t1 from the current time.

```{r}
Sys.time() - t1
```

The same line of thinking applies to addition and other compatison operators. If you want more control over the units when finding the above differnece in times you can use the difftime() which allows you to specify a **units** parameter.

```{r}
difftime(Sys.time(), t1, units = 'days')
```