---
title: "Week 1 - R Basics - File Operations"
author: "Sally Longmore"
date: "21/01/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## File Operations

Get the current working directory

```{r}
getwd()
```

List the objects in the current workspace

```{r}
ls()
```

List all the files in your current working directory

```{r}
list.files()
```

or you can use dir()

```{r}
dir()
```

You can use the args() function to determine the arguments to a function, for example

```{r}
args(list.files)
```

You can use dir.create() to create a new directory

```{r}
dir.create("testdir")
```

You can set the working directory using setwd()

```{r}
setwd("testdir")
```

You can create files using file.create()

```{r}
file.create("mytest.R")
```

You can check if a file exists by using file.exists()

```{r}
file.exists("mytest.R")
```

You can check the file information by using file.info()

```{r}
file.info("mytest.R")
```

You can also use the $ operator to grab specific items, for example the mode

```{r}
file.info("mytest.R")$mode
```

You can rename files using file.rename()

```{r}
file.rename("mytest.R", "mytest2.R")
```

You can copy a file by using file.copy()

```{r}
file.copy("mytest2.R", "mytest3.R")
```


You can delete files using file.remove()

```{r}
file.remove("mytest3.R")
```

You can get the relative path using file.path()

```{r}
file.path("mytest3.R")
```

You can pass folders into path to generate OS independent paths

```{r}
file.path("folder1", "folder2")
```

You can use dir.create() to make folders, but it will fail if parent folders do not exist.  
To overcome this, you can set the argument **recusrive** to **TRUE** and it will create the entire folder structure.

```{r}
dir.create(file.path("testdir2", "testdir3"), recursive = TRUE)
```

Go back to the original working directory we stored in the variable old.dir

```{r}
setwd(old.dir)
```


