---
title: "Week 1- R Basics - Subsetting Vectors"
author: "Sally Longmore"
date: "22/01/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this lesson we will see how to extract elements from a vector based on some conditions that we specify.

For example, we may only be interested in the first 20 elements of a vector, or only the elements that are not NA, or only those that are positive or correspond to a specific variable of interest. By the end of this lesson, you will know how to handle each of these scenarios.

A vector call x has been created that contains a random ordering of 20 numers from a standard normal distribution and 20 NAs. Type x now to see what it looks like.

```{r}
x
```

The way you tell R that you want to select some particular elements (i.e. a 'subset') from a vector is by placing an 'index vector' in square brackets immediately following the name of the vector.

For example:

```{r}
x[1:10]
```

Index vectors come in four different flavours, logical vectors, vectors of positive integers, vectors of negative integers, and vectors of character strings -- each of which we'll cover in this lesson.

Let's start by indexing with logical vectors. One common scenario when working with real-world data is that we want to extract all elements of a vector that are not NA (i.e. missing data). Recall that is.na(x) yields a vector of logical values the same length as x, with TRUEs corresponding to NA values in x and FALSEs corresponding to non-NA values in x.

What do you think x[is.na(x)] will give you?

```{r}
x[is.na(x)]
```

It will give you a vector of all NAs

Recall that **!** is the logical expression NOT, so !is.na(x) can be read as **is not NA**. Therefore, if we want to create a vector called y that contains all of the non-NA values from x, we can use y <- x[!is.na(x)]. Give it a try.

```{r}
y <- x[!is.na(x)]
y
```

Now we have a vector **y** that has isoloated the non-missing values of x, we can subset y as we please.  

Recall that the expression y > 0 will give us a vector of logical values the same length as y, with TRUEs corresponding to values of y that are greater than zero and FALSEs corresponding to values of y that are less than or equal to zero. What do you think y[y > 0] will give you?

```{r}
y[y > 0]
```

A vector of all the positive elements of y.

We could not use x[x > 0] because we would get a vector with elements containing including NAs as well as elements > 0.

Combining our knowledge of logical operators with our new knowledge of subsetting, we could do this.

```{r}
x[!is.na(x) & x > 0]
```

In this case we request only the values of x that are both not NA and greater than zero.

We have already looked at subsetting just the first 10 values **x[1:10]**. In this case we're providing a vector of positive integers inside of the square brackets, which tells R to return only the elements numbered 1 through 10.

Many programming languages what's called 'zero-based indexing' where the first element is considered element 0. R uses 'one-based indexing' where the first element is 1.

Can you figure out how we would subset the 3rd, 5th and 7th elements of x?

```{r}
x[c(3, 5, 7)]
```

It is important when using integer vectors to subset our vector x, we stick with the set of indexes {1, 2, ..., 40} since x only has 40 elements. What happens if we ask for the zero element of x?

```{r}
x[0]
```

As you might expect, we get nothing useful.

R doesn't prevent us from doing this or asking for the 3000th element of a vector containing 40 elements.

```{r}
x[3000]
```

This time you get NA, because it doesn't exist.

What if we are interested in all elements except the 2nd and 10th?

We can do this by using the -2 and -10 in the element list.

```{r}
x[c(-2, -10)]
```

A shorthand way of specifying multiple negative numbers is to put a negative sign out the front of the vector of positive numbers.

```{r}
x[-c(2, 10)]
```

So far we have covered three types of index vectors, logical, positive integer and negative integer. The only remaining type requires us to introduce the concept of *named* elements.

Create a numeric vector with three named elements using vect <- c(foo = 11, bar = 2, norf = NA).

```{r}
vect <- c(foo = 11, bar = 2, norf = NA)
```

When we print vect to the console, you will see that each element has a name.

```{r}
vect
```

We can also get the names of vect by passing vect as an argument to the names() function.

```{r}
names(vect)
```

Alternatively, we can create an unnamed vector vect2 with c(11, 2, NA)

```{r}
vect2 <- c(11, 2, NA)
```

Then we can add the *names* attribute to vect2 after the fact with names(vect2) <- c("foo", "bar", "norf").

```{r}
names(vect2) <- c("foo", "bar", "norf")
```

Now, let's check that vect and vect2 are the same by passing them as arguments to the identical() function.

```{r}
identical(vect, vect2)
```

Indeed, vect and vect2 are identical named vectors.

With named vectors, we can select a vector using the name of the vector.

```{r}
vect["bar"]
```

Likewise, we can specify a vector names with vect[c("foo", "bar")]

```{r}
vect[c("foo", "bar")]
```

Now you know all four methods of subsetting data from vectors. Different approaches are best in different scenarios and when in doubt, try it out!