---
title: "Week 1 - R Basics - Sequences of Numbers"
author: "Sally Longmore"
date: "22/01/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The simplest way to create a sequence of numbers in R is by using the colon ":" operator.

```{r}
1:20
```
We can also use it to create a sequence of real numbers, for example pi:10

```{r}
pi:10
```

The result is a vector of real numbers starting with pi (3.142...) and increasing in increments of 1.  
The upper limit of 10 is never reached, since the next number in our sequence would be greater than 10.

What happens if we do 15:1?

```{r}
15:1
```

It creates a vector from 15 to 1 in increments of 1

To get help on an operator such as ":", you must enclose the operator in back tick like this **`:`**

If you want more control creating a sequence of numbers you can use the seq() function. The most basic operation of the seq() function is the same as that of the ":" operator

```{r}
seq(1, 20)
```

This gives us a sequence between 1 and 20 with intervals of 1

No if we want to change the interval between numbers in the sequence, we can use seq() to do that

```{r}
seq(0, 10, by=0.5)
```

Lets say we want a sequence of 30 numbers equally spaced between 5 and 10, we can do this with seq()

```{r}
my_seq <- seq(5, 10, length=30)
my_seq
```

To confirm the lenght of my_seq use the lenght() function

```{r}
length(my_seq)
```

Lets say we want to generate a sequence of integers from 1 to N, where N represetns the length of my_seq vector

'
```{r}
1:length(my_seq)
```

Or we can use seq()

```{r}
seq(along.with = my_seq)
```

R also has a built in function for this purpose as well called seq_along()

```{r}
seq_along(my_seq)
```

One more function for creating a sequence of numbers if rep(), which stands for **replicate**.

If we want to create a vector that contains 40 zeros, we can use

```{r}
rep(0, times = 40)
```

If instead we want to our vector to contain 10 repetitions of the vector (0, 1, 2) we can use a vector instead of the number.

```{r}
rep(c(0, 1, 2), times = 10)
```

What if we want a vector that contains 10 zeros, 10 ones and 10 twos. We can do this with the each argument instead of times.

```{r}
rep(c(0, 1, 2), each = 10)
```
