---
title: "Week 1 - QUiz"
author: "Sally Longmore"
date: "19/01/2020"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Qustion 1 - 10
### 1. (Ver A) The R language is a dialect of which of the following programming languages?

- S

### 1. (Ver B) R was developed by statisticians working at
- The University of Auckland

### 2. The definition of free software consist of four freedoms (freedoms 0 through 3). Which of the following is not one of the freedoms that are part of the definition? Select all that apply
- **The freedom to prevent users from using the software for undesirable purposes**
- **The freedom to restrict access the source code**
- The freedom to run the program for any purpose
- The freedom to redistribute copies so you can help your neighbour
- The freedom to improve the program, and release your improvements to the public, so that the whole community benefits
- The freedom to study how the program works, and adapt it to your needs
- **The freedom to sell the software for any price.**

### 3. In R the following are all atomic data types EXCEPT: (Select all that apply)
- **array**
- numeric
- **data frame**
- **martrix**
- integer
- complex
- **list**
- **table**
- character
- logical

### 4. (Ver A) If I execute the expression x <- 4L in R, what is the class of the object 'x' as determined by the class() function?
- integer  
*the 'L' suffix creates an integer vector as opposed to a numeric vector*

```{r}
x <- 4L;
class(x)
```

### 4. (Ver B) If I execute the expression x <- 4 in R, what is the class of the object 'x' as determined by the class() function?
- numeric  

```{r}
x <- 4;
class(x)
```

### 5. (Ver A) What is the class of the object defined by the expression x <- c(4, "a", TRUE)?
- character  
*The character class is the "lowest common denominator" here and so all elements will be coerced into that class*

```{r}
x <- c(4, "a", TRUE);
class(x)
```

### 5. (Ver B) What is the class of the object defined by the expression x <- c(4, TRUE)?
- numeric  
*The numeric class is the "lowest common denominator" here and so all elements will be coerced into that class*

```{r}
x <- c(4, TRUE);
class(x)
```


### 6. (Ver A) If I have two vectors x <- c(1, 3, 4) and y <- c(3, 2, 10), what is produced by the expression rbind(x, y)?
- a matrix with two rows and three columns  
*The 'rbind' function treats vectors as if they were rows of a matrix. It then takes those vectors and binds them together row-wise to create a matrix*

```{r}
x <- c(1, 3, 4);
y <- c(3, 2, 10);
rbind(x, y)
```

### 6. (Ver B) If I have two vectors x <- c(1, 3, 5) and y <- c(3, 2, 10), what is produced by the expression cbind(x, y)?
- a matrix with two2 columns and 3 rows  
*The 'cbind' function treats vectors as if they were columns of a matrix. It then takes those vectors and binds them together column-wise to create a matrix*

```{r}
x <- c(1, 3, 5);
y <- c(3, 2, 10);
cbind(x, y)
```


### 7. A key property of vectors in R is that
- elements of a vector all must be the same class

### 8. (Ver A) Suppose I have a list defined as x <- list(2, "a", "b", TRUE). What does x[[2]] give me? Select all that apply.
- a character vector of length 1
- a character vector containing the letter "a"

```{r}
x <- list(2, "a", "b", TRUE);
x[[2]]
class(x[[2]])
```

### 8. (Ver B) Suppose I have a list defined as x <- list(2, "a", "b", TRUE). What does x[[1]] give me? Select all that apply.
- a numeric vector of length 1
- a numeric vector containing the element 2

```{r}
x <- list(2, "a", "b", TRUE);
x[[1]]
class(x[[1]])
```

### 9. Suppose I have a vector x <- 1:4 and a vector y <- 2. What is produced by the expression x + y?
- A numeric vector with elements 3, 4, 5, 6

```{r}
x <- 1:4;
y <- 2;
z = x + y
z
class(z)
```

### 10. Suppose I have a vector x <- c(3, 5, 1, 10, 12, 6) and I want to set all elements of this vector that are less than 6 to be equal to zero. What R code achieves this? Select all that apply.
- x[x <= 5] <- 0
*You can create a logical vector with the expression x <= 5 and then use the [ operator to subset the original vector x*

```{r}
x <- c(3, 5, 1, 10, 12, 6);
x[x <= 5] <- 0;
x
```

- x[x %in% 1:5] <- 0  
*You can create a logical vector with the expression x %in% 1:5 and then use the [ operator to subset the original vector x*

```{r}
x <- c(3, 5, 1, 10, 12, 6);
x[x %in% 1:5] <- 0;
x
```

- x[x < 6] <- 0  
*you can create a logical vector with the expression x < 6 and then use the [ operator to subset the original vector x*

```{r}
x <- c(3, 5, 1, 10, 12, 6);
x[x < 6] <- 0;
x
```

## Question 11 - 20
### For questions 11 - 20 use the provided dataset

Load the dataset first
```{r}
week1 <- read.csv("hw1_data.csv")
```

### 11. In the dataset provided for this quiz, what are the columns for the dataset?

We need to extract the header info using **names()**

```{r}
names(week1)
```

### 12. Extract the first 2 rows of the data frame and print them to the console. What does the output look like?

We need to use the **[** operator to get a subset of the data frame

```{r}
week1[1:2, ]
```

### 13. How many observations (i.e. rows) are in this dataframe?

We need to use the **nrow()** function to compute the number of rows in the data frame

```{r}
nrow(week1)
```

### 14. Extract the last 2 tows of the data frame and print them to the console

We can use the **tail()** function to select rows from the end of the data frame working backward

```{r}
tail(week1, 2)
```

### 15. What is the value of Ozone in the 47th row?

We can use the **[** operator to select a subset of the data frame

```{r}
week1[47, "Ozone"]
```

### 16. How many missing values are in the Ozone column of this data frame?

We can use the **is.na()** function to test for missing values.
We can then remove the FALSE values using na.rm 
```{r}
bad = is.na(week1[, "Ozone"]);
sum(bad, na.rm = TRUE)
```

### 17. What is the mean of the Ozone column in this data frame? Excluding missing values (coded as NA) from this calculation?

We fist need to remove the missing values, we can do this using the **complete.cases** function, we need to also use the **[** operator here to insure that we only remove rows where the NA value is in the *Ozone* column.
We then need to use the **mean** command to calculate the mean. In that command we need to ensure that we are only including the rows that have complete data, and that we only include the *Ozone* column.

```{r}
good <- complete.cases(week1[, "Ozone"]);
mean(week1[good, "Ozone"])
```

### 18. Extract the subset of rows of the data frame where Ozone values are above 31 and Temp values are above 90. What is the mean of Solar.R in this subset?

See in code comments.

```{r}
Ozone <- 31;
Temp <- 90;
## First select only the Ozone and Temp columns
sel <- week1[, c("Ozone", "Temp", "Solar.R")];
## Remove NA's
selComp <- complete.cases(sel);
sel <- sel[selComp, ];
## Select rows where Ozone > 31
selOzone <- sel["Ozone"] > Ozone;
sel <- sel[selOzone, ];
## Select rows where Temp > 90
selTemp <- sel["Temp"] > Temp;
sel <- sel[selTemp, ];
## Get the mean of Solar.R from the selected rows
mean(sel[, "Solar.R"])
```

### 19. What is the mean of "Temp" when "Month" is equal 6.

```{r}
selMonth <- 6;
## First select only the rows that are Month 6
sel <- week1["Month"] == selMonth;
sel <- week1[sel, ];
## Get the mean of Temp
mean(sel[, "Temp"])
```

### 20. What was the maximum ozone value in the month of May (i.e. Month is equal to 5)?

```{r}
selMonth <- 5;
## First select only the rows that are Month 6
sel <- week1["Month"] == selMonth;
sel <- week1[sel, "Ozone"];
## Remove NA's
selComp <- complete.cases(sel);
sel <- sel[selComp];
## Get the max value of Ozone
max(sel)
```

